## Exploring the U.S. NOAA storm database to analysis of severe weather events that are most harmful to the population health and economy.

### Reproducible Research: Peer Assessment 2
#### J.A (newja1390@gmail.com)


### 1. Synopsis

The basic goal of this report is to explore the U.S. National Oceanic and Atmospheric Administration's (NOAA) Storm Database and answer these basic questions about severe weather events:

1. which types of events are most harmful with respect to population health?
2. which types of events have the greatest economic consequences?

The events in the database start in the year 1950 and end in November 2011.
The data of fatalities, injuries, property and crop damage will be used to decide which types of events are most harmful to the population health and economy.

### 2. Data Processing

##### 2.1. Load libraries
Load necessary libraries to perform data analysis.
```{r Loadlibraries, warning=FALSE, message=FALSE}
library(utils)
library (stringr)
library(plyr) 
library(ggplot2) 
library(grid)
library(gridExtra)
```

##### 2.2. Loading data
Unzip the data source file and load the data set.
```{r LoadData, warning=FALSE, message=FALSE}
 
 # unzip data file and read data from dataset
   zipfile="repdata_data_StormData.csv.bz2"
   if(file.exists(zipfile))
   {
        unzip(zipfile,overwrite = TRUE)
        dataSet <- read.csv("repdata_data_StormData.csv")
    }
```

##### 2.3. Preprocessing the data
```{r PreprocessingData, cache = TRUE}

 # Select needed colums of dataset for this analysis:
selectedColumns <- c("BGN_DATE", "EVTYPE", "FATALITIES", "INJURIES",
                     "PROPDMG","PROPDMGEXP", "CROPDMG", "CROPDMGEXP")
 
stormData <- dataSet[, selectedColumns]

# extract the year of events and store it in a new column('year'), in storm dataset.
stormData$year <- as.numeric(format(as.Date(stormData$BGN_DATE,
                                            format = "%m/%d/%Y %H:%M:%S"), "%Y"))

# Plotting the histogram of the events in years 1950-2011.
hist(stormData$year, 
     main="Histogram of the total events occured each year",  
     sub="",
     xlab="Year (1950-2011)",
     ylab="Events Frequency",
     col="red",
     breaks = 20)
```


Based on code book description([Storm Events][1]), data for years prior to 1996 is incomplete and might be incorrect. So we will filter data for years prior to 1996.

```{r , cache = TRUE}
# Filter data for years prior to 1996 
stormData <- stormData[stormData$year >= 1996, ]

# Clean event types
stormData$EVTYPE <- str_trim(str_to_upper(stormData$EVTYPE))
stormData$EVTYPE <-str_replace(stormData$EVTYPE,'[ ]+', " ")
stormData <- stormData[str_detect(stormData$EVTYPE, "^SUMMARY")== FALSE,]

# coefficients of property and crop damages values before cleaning :
#
# unique(stormData$PROPDMGEXP)
## [1] K   M B 0
## Levels:  - ? + 0 1 2 3 4 5 6 7 8 B h H K m M

# unique(stormData$CROPDMGEXP)
## [1] K   M B
## Levels:  ? 0 2 B k K m M

# Clean the coefficients of property and crop damages values in the PROPDMGEXP and CROPDMGEXP columns.
stormData$PROPDMGEXP <- as.character(stormData$PROPDMGEXP)
stormData$PROPDMGEXP <- gsub("h|H", "2", stormData$PROPDMGEXP)
stormData$PROPDMGEXP <- gsub("k|K", "3", stormData$PROPDMGEXP)
stormData$PROPDMGEXP <- gsub("m|M", "6", stormData$PROPDMGEXP)
stormData$PROPDMGEXP <- gsub("B", "9", stormData$PROPDMGEXP)
stormData$PROPDMGEXP <- gsub("\\-|\\+|\\?", "0", stormData$PROPDMGEXP)
stormData$PROPDMGEXP <- as.numeric(stormData$PROPDMGEXP)
stormData$PROPDMGEXP[is.na(stormData$PROPDMGEXP)] = 0

stormData$CROPDMGEXP <- as.character(stormData$CROPDMGEXP)
stormData$CROPDMGEXP <- gsub("h|H", "2", stormData$CROPDMGEXP)
stormData$CROPDMGEXP <- gsub("k|K", "3", stormData$CROPDMGEXP)
stormData$CROPDMGEXP <- gsub("m|M", "6", stormData$CROPDMGEXP)
stormData$CROPDMGEXP <- gsub("B", "9", stormData$CROPDMGEXP)
stormData$CROPDMGEXP <- gsub("\\-|\\+|\\?", "0", stormData$CROPDMGEXP)
stormData$CROPDMGEXP <- as.numeric(stormData$CROPDMGEXP)
stormData$CROPDMGEXP[is.na(stormData$CROPDMGEXP)] = 0

# coefficients of property and crop damages values after cleaning :
#
# unique(stormData$PROPDMGEXP)
## [1] 3 0 6 9

# unique(stormData$CROPDMGEXP)
## [1] 3 0 6 9

# transform data set and calculate properties and crops damages amount in dollars.
stormData <- mutate(stormData, 
                    PROPERTIES = (PROPDMG * 10^PROPDMGEXP), 
                    CROPS = (CROPDMG * 10^CROPDMGEXP))

```

##### 2.4. Find events that are most harmful with respect to population health
Calculate the number of fatalities and injuries that are caused by the severe weather events.
Then get the first 20 most severe types of weather events.

```{r PopulationHealthHarmfulEvents, cache = TRUE}

# group fatalities by event type and get sum of fatalities for each type
fatalities <- aggregate(x=list(FATALITIES=stormData$FATALITIES),
                               by=list(EVTYPE=stormData$EVTYPE),
                               FUN=sum, na.rm=TRUE)

# group injuries by event type and get sum of injuries for each type
injuries <- aggregate(x=list(INJURIES=stormData$INJURIES),
                               by=list(EVTYPE=stormData$EVTYPE),
                               FUN=sum, na.rm=TRUE)

# sort by event frequency and select top 20 Injuries and Fatalities to compare.
Top20Fatalities <- head(fatalities[with(fatalities, order(-FATALITIES)), ], 20)
Top20Injuries   <- head(injuries[with(injuries, order(-INJURIES)), ], 20)

# reset row names to ordered numbers 1:20
row.names(Top20Fatalities) <- seq(length=nrow(Top20Fatalities))
row.names(Top20Injuries) <- seq(length=nrow(Top20Injuries))

```

##### 2.5. Find events that have greatest economic consequences
There are two different factors for economic consequences, those causing property damage and those causing crop damage.

```{r EconomicConsequencesEvents, cache = TRUE}
# group property damages by event type and get sum of property damages for each type
properties <- aggregate(x=list(PROPERTIES=stormData$PROPERTIES),
                               by=list(EVTYPE=stormData$EVTYPE),
                               FUN=sum, na.rm=TRUE)

# group crops damages by event type and get sum of crops damage for each type
crops <- aggregate(x=list(CROPS=stormData$CROPS),
                               by=list(EVTYPE=stormData$EVTYPE),
                               FUN=sum, na.rm=TRUE)

# sort by event frequency and select top 20 Property and Crop damages to compare.
Top20Properties <- head(properties[with(properties, order(-PROPERTIES)), ], 20)
Top20Crops <- head(crops[with(crops, order(-CROPS)), ], 20)

# reset row names to ordered numbers 1:20
row.names(Top20Properties) <- seq(length=nrow(Top20Properties))
row.names(Top20Crops) <- seq(length=nrow(Top20Crops))

```

### 3. Results
##### 3.1. Show the top 20 Injuries and  Fatalities and plot Injuries vs. Fatalities
```{r  TopHealthDamages}
#show top 20 Injuries and  Fatalities data
Top20Fatalities
Top20Injuries
```

```{r InjuriesVsFatalities, fig.width=10, warning=FALSE}
# need to specify the levels of factor in the approprate order 
Top20Fatalities <- within(Top20Fatalities, 
                          EVTYPE <- factor(x = EVTYPE, levels = Top20Fatalities$EVTYPE))

fatalitiesPlot <- qplot(EVTYPE, data=Top20Fatalities,weight=FATALITIES, geom="bar",binwidth=1) +               
                        theme(axis.text.x = element_text(angle = 45,hjust = 1)) +
                        xlab("Severe Weather Type") +
                        ylab("Number of Fatalities") +
                        ggtitle("Total Fatalities\n by Severe Weather Events from 1996-2011")

Top20Injuries <- within(Top20Injuries, 
                        EVTYPE <- factor(x = EVTYPE, levels = Top20Injuries$EVTYPE))

injuriesPlot <- qplot(EVTYPE, data=Top20Injuries,weight=INJURIES, geom="bar", binwidth=1) +      
                      theme(axis.text.x = element_text(angle = 45,hjust = 1)) +
                      xlab("Severe Weather Type") + 
                      ylab("Number of Injuries") +
                      ggtitle("Total Injuries\n by Severe Weather Events from 1996-2011")

# add an 1 by 2 rectangular grid to existing plots
grid.arrange(fatalitiesPlot, injuriesPlot, ncol = 2)

```

##### 3.2. Show the top 20 Properties and Crops damages and plot Properties vs. Crops
```{r  TopEconomicDamages}
#show top 20 Properties and  Corps damages data
Top20Properties
Top20Crops
```

```{r PropertiesVsCrops, fig.width=10, warning=FALSE}
# need to specify the levels of factor in the approprate order 
Top20Properties <- within(Top20Properties, 
                          EVTYPE <- factor(x = EVTYPE, levels = Top20Properties$EVTYPE))

propertiesPlot <- qplot(EVTYPE, data=Top20Properties,weight=PROPERTIES, geom="bar",binwidth=1) +               
                        theme(axis.text.x = element_text(angle = 45,hjust = 1)) +
                        xlab("Severe Weather Type") +
                        ylab("Properties Damages ($)") +
                        ggtitle("Total Properties\n by Severe Weather Events from 1996-2011")

Top20Crops <- within(Top20Crops, 
                        EVTYPE <- factor(x = EVTYPE, levels = Top20Crops$EVTYPE))

cropsPlot <- qplot(EVTYPE, data=Top20Crops,weight=CROPS, geom="bar", binwidth=1) +      
                      theme(axis.text.x = element_text(angle = 45,hjust = 1)) +
                      xlab("Severe Weather Type") + 
                      ylab("Crops Damages ($)") +
                      ggtitle("Total Crops\n by Severe Weather Events from 1996-2011")

# add an 1 by 2 rectangular grid to existing plots
grid.arrange(propertiesPlot, cropsPlot, ncol = 2)

```

### 3. Conclosion
Excessive heat, tornado and flood are most harmful events with respect to population health, while flood, drought and hurricane/typhoon have the greatest economic impacts.

--------------------------------------------------------------------------------
[1]: http://ire.org/nicar/database-library/databases/storm-events/

 